---
title: 'Homework: iterative method'
author: "Leon Di Stefano"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
header-includes:
    - \newcommand{\iidsim}{\overset{\mathrm{ind}}{\sim}}
    - \newcommand{\d}{\mathrm{d}\;}
    - \newcommand{\E}{\mathrm{E}}
    - \newcommand{\Var}{\mathrm{Var}}
    - \newcommand{\Cov}{\mathrm{Cov}}
    - \newcommand{\P}{\mathrm{P}}
    - \renewcommand\vec{\boldsymbol}
---

```{r setup, include = FALSE}
# source(file.path("..", "R", "util.R"))
# required_packages <- c('RSpectra')
# install_and_load_packages(required_packages)
require('RSpectra')
knitr::opts_chunk$set(
  # eval = FALSE # For debugging
  )

require(tidyverse)
require(cowplot)
require(lemon)
theme_set(theme_cowplot())
theme_update(
  strip.background = element_blank(),
  strip.text.y = element_text(angle = 0),
  )
```

# Problem 1
Auto regressive processes can be viewed as a discrete analog of Ornstein–Uhlenbeck process &mdash; which coincides with Gaussian process based on an exponential covariance matrix &mdash; and hence is an example of Gaussian Markov random fields.
For instance, stationary lag-1 auto-regressive process 
$$x_t = \phi x_{t - 1} + \sqrt{1 - \phi^2} \, \epsilon_t, 
  \quad x_0 \sim \mathcal{N}(0, 1), 
  \quad \epsilon_t \mathbin{\overset{\small \textrm{i.i.d.}}{\sim}} \mathcal{N}(0, 1)$$
has the _tri-diagonal_ precision matrix
$$\boldsymbol{\Sigma}^{-1} = \frac{1}{1 - \phi^2} 
  \begin{bmatrix} 
  1 & -\phi & 0 & & & \ldots & 0 \\
  -\phi & 1 + \phi^2 & -\phi & 0 & & & \vdots \\
  0 & -\phi & 1 + \phi^2 & -\phi & 0 & & \\
    & & \ddots & \ddots & \ddots & & \\
    & & 0 & -\phi & 1 + \phi^2 & -\phi & 0 \\
  \vdots & &   & 0 & -\phi & 1 + \phi^2 & -\phi \\
  0 & \ldots &   &   & 0 & -\phi & 1\\
  \end{bmatrix}.$$
More generally, a lag-$k$ (non-stationary) auto-regressive process has a _banded_ precision matrix with bandwidth $k$.

Implement a fast matrix-vector $\boldsymbol{v} \to \boldsymbol{\Sigma}^{-1} \boldsymbol{v}$ operation, exploiting the structure of the AR-1 precision matrix.
Then use this function to find the top 10 principal components of $\boldsymbol{\Sigma}$ (not $\boldsymbol{\Sigma}^{-1}$) via Lanczos algorithm provided via `RSpectra::eigs_sym`.

> Let
> \begin{align}
a &= \frac{1}{1 - \phi^2} \\
b &= 1 + \phi^2 \\
c &= -\phi
> \end{align}
> and $\vec v$ have elements $v_1, \ldots, v_T$.
>
> Then $\vec w = \vec \Sigma^{-1} \vec v$ is given by $a$ times
$$
\begin{bmatrix} 
v_1 + c v_2 \\
b v_2 + c(v_1 + v_3) \\
\vdots \\
b v_{T - 1} + c(v_{T - 2} + v_T) \\
v_T + c v_{T - 1}
\end{bmatrix}
=
\begin{bmatrix} 
v_1 \\
b \vec v_{2:T-1} \\
v_T
\end{bmatrix}
+
c(\mathrm{lead}(\vec v) + \mathrm{lag}(\vec v))
$$
> where the out-of-bounds indices in "lead" and "lag" are set to 0.
>
> I believe a vectorized, numerically sensible way to compute this is to first compute $b\vec v$, then manually mutate the first and last elements in-place, then add the second term, then scale the whole thing by $a$. There are fast implementations of lead and lag in `dplyr`.

```{r}
ar_length <- 4096
auto_corr <- .9 # Corresponds to `\phi` above

ar_precision_matvec <- function(v, auto_corr) {
  # Fill in: note that you can vectorize the calculation and do *not* need a for-loop. 
  # (Hint: how would you efficiently carry out a matrix-vector operation if the matrix has non-zero entries only along the sub or super diagonal?)
  
  a <- 1/(1 - auto_corr^2)
  b <- 1 + auto_corr^2
  c <- -auto_corr
  
  w <- b * v
  w[1]         <- v[1]
  w[length(v)] <- v[length(v)]
  w <- w + c * (
    lead(v, default = 0) + 
      lag(v, default = 0))
  w <- a*w
  
  return(w)
}

ar_eig <- eigs_sym(
  ar_precision_matvec, args = auto_corr,
  # Fill in
  n = ar_length,
  k = 10,
  which = "SM", # Largest eigs of cov => smallest of prec
  opts = list(
    ncv = 100, # Spectrum distribution of AR-1 process is not very spread out on the extreme ends and is actually a hard case for Lanczos. So it helps to have more Lanczos vectors than the default for faster convergence.
    maxitr = 10^3, # Cap it just in case
    retvec = TRUE # More efficient to do without eigenvectors when not needed
  ),
)
```

Now, directly compute the eigen decomposition of $\boldsymbol{\Sigma}$ (not $\boldsymbol{\Sigma}^{-1}$) and compare its output with the principal components and associated variances found via Lancsoz algorithm.

> With the parametrization given, 
$$
\Var(x_t) = \phi^2 \Var(x_{t-1}) + (1 - \phi^2)
$$
so that all of the terms have variance 1, and
$$
\Cov(x_t, x_{t-k}) = \phi \Cov(x_{t - 1}, x_{t - k})
$$
so that $\Cov(x_t, x_{t-k}) = \phi^{k}$.
The covariance matrix thus has entries
$$ \sigma_{ij} = \phi^{|i - j|}.$$

```{r}
ar_cov_matrix <-
  matrix(
    auto_corr^(abs(
      rep(1:ar_length, ar_length) - rep(1:ar_length, each = ar_length))),
    nrow = ar_length)

ar_cov_eig_direct <-
  eigen(ar_cov_matrix)
```

> Comparing the two: first, eigenvalues:

```{r}
ggplot2::qplot(
  rev(1/ar_eig$values),
  ar_cov_eig_direct$values[1:10],
  size = I(.5)
) +
  theme(aspect.ratio = 1) +
  geom_abline(color = "orange", alpha = .3) +
  xlab("Lanczos of precision matrix") +
  ylab("Direct computation") +
  ggtitle("Top 10 eigenvalues, computed two ways")
```
> Looks good.
>
> Next, vectors. Since the eigenvectors are only determined up to sign, I'll compare them using absolute correlations, which correspond to cosines of angles between the corresponding directions.

```{r}
sqrt(colSums(
  (ar_eig$vectors - ar_cov_eig_direct$vectors[,10:1])^2))
```

> This looks right: the eigenvectors are given as unit vectors and are only determined up to sign. The zeroes correspond to same direction, same orientation, and the 2s correspond to same direction, opposite orientation.

**Remark:** 
For banded matrices, there actually are even more efficient approaches.
To get a sense of special routines available for banded matrices, you can take a look at `*_banded` functions in [SciPy's linear algebra routines](https://docs.scipy.org/doc/scipy/reference/linalg.html).
Even those functions represent only a subset of available numerical linear algebra techniques; see
[LAPACK documentation for SVD](https://www.netlib.org/lapack/lug/node32.html) for example.
